---
title: "PeMS 3, modeling"
date: "Last modified: `r format(Sys.time(), '%d-%m-%Y.')`"
output:
  html_document:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
    highlight: pygments
    theme: flatly
    code_folding: hide # class.source = "fold-hide" to hide code and add a button to show it
    df_print: paged
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    fig_caption: true
    code_download: true
    css: visual.css
always_allow_html: true
bibliography: 
  - references.bib
  - grateful-refs.bib
header-includes:
  - \newcommand{\ar}{\mathbb{R}}
  - \newcommand{\llav}[1]{\left\{#1\right\}}
  - \newcommand{\pare}[1]{\left(#1\right)}
  - \newcommand{\Ncal}{\mathcal{N}}
  - \newcommand{\Vcal}{\mathcal{V}}
  - \newcommand{\Ecal}{\mathcal{E}}
  - \newcommand{\Wcal}{\mathcal{W}}
  - \newcommand{\almosteverywhere}{\mathrm{a.e.}\;}
---

Go back to the [Contents](about.html) page.

<div style="color: #2c3e50; text-align: right;">
********  
<strong>Press Show to reveal the code chunks.</strong>  

********
</div>



Go back to the [About page](about.html). 


This vignette compares different models for PeMS data. It uses [**`pems_repl1_data.RData`**](https://github.com/leninrafaelrierasegura/GWMF/blob/main/data_files/pems_repl1_data.RData), which is a file with a graph and data created in [pems_repl1.html](pems_repl1.html).

Let us set some global options for all code chunks in this document.


```{r}
# Create a clipboard button on the rendered HTML page
source(here::here("clipboard.R")); clipboard
# Set seed for reproducibility
set.seed(1938) 
# Set global options for all code chunks
knitr::opts_chunk$set(
  # Disable messages printed by R code chunks
  message = FALSE,    
  # Disable warnings printed by R code chunks
  warning = FALSE,    
  # Show R code within code chunks in output
  echo = TRUE,        
  # Include both R code and its results in output
  include = TRUE,     
  # Evaluate R code chunks
  eval = FALSE,       
  # Enable caching of R code chunks for faster rendering
  cache = FALSE,      
  # Align figures in the center of the output
  fig.align = "center",
  # Enable retina display for high-resolution figures
  retina = 2,
  # Show errors in the output instead of stopping rendering
  error = TRUE,
  # Do not collapse code and output into a single block
  collapse = FALSE
)
# Start the figure counter
fig_count <- 0
# Define the captioner function
captioner <- function(caption) {
  fig_count <<- fig_count + 1
  paste0("Figure ", fig_count, ": ", caption)
}


# Define the function to truncate a number to two decimal places
truncate_to_two <- function(x) {
  truncated <- floor(x * 100) / 100
  sprintf("%.2f", truncated)
}
```

Below we load the necessary libraries.

```{r, eval = TRUE}

# inla.upgrade(testing = TRUE)
# remotes::install_github("inlabru-org/inlabru", ref = "devel")
# remotes::install_github("davidbolin/rspde", ref = "devel")
# remotes::install_github("davidbolin/metricgraph", ref = "devel")
library(INLA)
library(inlabru)
library(rSPDE)
library(MetricGraph)

library(dplyr)
library(plotly)
library(scales)
library(patchwork)

library(ggplot2)
library(cowplot)
library(ggpubr) #annotate_figure()
library(grid) #textGrob()
library(ggmap)

library(viridis)
library(OpenStreetMap)


library(tidyr)
library(sf)

library(here)
library(rmarkdown)
library(grateful) # Cite all loaded packages
```


```{r}
source_if_updated <- function(name) {
  rmd_file  <- here::here(paste0(name, ".Rmd"))
  hash_file <- here::here(paste0(name, ".hash"))
  r_file    <- here::here(paste0(name, ".R"))
  purl_file <- here::here("old/purl_log.txt")
  
  new_hash <- digest::digest(file = rmd_file, algo = "md5")
  
  old_hash <- if (file.exists(hash_file)) {
    readLines(hash_file, warn = FALSE)
  } else {
    ""
  }
  
  if (!identical(new_hash, old_hash)) {
    capture.output(
      knitr::purl(rmd_file, output = r_file),
      file = purl_file
    )
    writeLines(new_hash, hash_file)
  }
  
  source(r_file)
}
```


```{r}
# capture.output(
#   knitr::purl(here::here("functionality1.Rmd"), output = here::here("functionality1.R")),
#   file = here::here("old/purl_log.txt")
# )
# source(here::here("functionality1.R"))
source_if_updated("functionality1")
```

Below we define function `process_model_results()` to extract the summary of the parameters of the model.



```{r}
process_model_results <- function(fit, model) {
  fit_spde <- rspde.result(fit, "field", model, parameterization = "spde")
  fit_matern <- rspde.result(fit, "field", model, parameterization = "matern")
  df_for_plot_spde <- gg_df(fit_spde)
  df_for_plot_matern <- gg_df(fit_matern)
  param_spde <- summary(fit_spde)
  param_matern <- summary(fit_matern)
  param_fixed <- fit$summary.fixed[,1:6]
  marginal.posterior.sigma_e = inla.tmarginal(
    fun = function(x) exp(-x/2), 
    marginal = fit[["internal.marginals.hyperpar"]][["Log precision for the Gaussian observations"]])
  quant.sigma_e <- capture.output({result_tmp <- inla.zmarginal(marginal.posterior.sigma_e)}, file = "/dev/null") 
  quant.sigma_e <- result_tmp
  statistics.sigma_e <- unlist(quant.sigma_e)[c(1,2,3,5,7)]
  mode.sigma_e <- inla.mmarginal(marginal.posterior.sigma_e)
  allparams <- rbind(param_fixed, param_spde, param_matern, c(statistics.sigma_e, mode.sigma_e))
  rownames(allparams)[nrow(allparams)] <- "sigma_e"
  return(list(allparams = allparams, df_for_plot_spde = df_for_plot_spde, df_for_plot_matern = df_for_plot_matern))
}
```

We first load the data in the file `pems_repl1_data.RData` and extract the data from the graph.

```{r}
# Load the data
load(here("data_files/new_pems_repl1_data.RData"))
# Extract the data from the graph
data <- graph$get_data()
```

Below we define the non-stationary parameters.

```{r}
# Non-stationary parameters
B.tau = cbind(0, 1, 0, cov, 0)
B.kappa = cbind(0, 0, 1, 0, cov)
```

We now model the speed records $y_i$ as 13 independent replicates satisfying
\begin{equation}
\label{applimodel}
    y_i|u(\cdot)\sim N(\beta_0 + \beta_1\text{mean.cov}(s_i) + u(s_i),\sigma_\epsilon^2),\;i = 1,\dots, 314,
\end{equation} 
where $u(\cdot)$ is a Gaussian process on the highway network. We consider stationary models with $\kappa,\tau>0$ and non-stationary models where $\tau$ and $\kappa$ are given by
\begin{equation}
\label{logregressions}
    \begin{aligned}
    \log(\tau(s)) &= \theta_1 + \theta_3 \text{std.cov}(s),\\
    \log(\kappa(s)) &= \theta_2 + \theta_4 \text{std.cov}(s).
\end{aligned}
\end{equation}

For each of the two classes of models, we consider three cases: when (1) $\nu$ is fixed to 0.5 or (2) 1.5, and (3) $\nu$ is estimated from the data. 

Below `cov` refers to $\text{std.cov}(s)$ and `mean_value` refers to $\text{mean.cov}(s)$.


```{r}
load(here::here("data_files/tau_from_graphlme.RData"))
log_tau_from_graphlme <- log(tau_from_graphlme)
log_kappa_from_graphlme <- log(kappa_from_graphlme)
```


```{r}
distance = seq(from = 0, to = 10, by = 0.1)
```

```{r}
# Define aux data frame to compute the distance matrix
aux <- data |> filter(repl == 1) |>
  rename(distance_on_edge = .distance_on_edge, edge_number = .edge_number) |> # Rename the variables (because graph$compute_geodist_PtE() requires so)
  as.data.frame() |> # Transform to a data frame (i.e., remove the metric_graph class)
  dplyr::select(edge_number, distance_on_edge)

# Compute the distance matrix
distmatrix <- graph$compute_geodist_PtE(PtE = aux,
                                             normalized = TRUE,
                                             include_vertices = FALSE)

# Compute the groups for one replicate
GROUPS <- list()
GROUPS2 <- list()
for (j in 1:length(distance)) {
  GROUPS[[j]] = list()
  GROUPS2[[j]] = list()
  for (i in 1:nrow(aux)) {
    GROUPS[[j]][[i]] <- which(as.vector(distmatrix[i, ]) <= distance[j])
    GROUPS2[[j]][[i]] <- which(as.vector(distmatrix[i, ]) > distance[j])
  }
}
# Compute the groups for all replicates, based on the groups of the first replicate
nrowY <- length(unique(data$repl))
ncolY <- nrow(filter(data, repl == 1))

NEW_GROUPS <- list()
for (j in 1:length(distance)) {
  my_list <- GROUPS[[j]]
  aux_list <- list()
  for (i in 0:(nrowY - 1)) {
  added_vectors <- lapply(my_list, function(vec) vec + i*ncolY)
  aux_list <- c(aux_list, added_vectors)
  }
  NEW_GROUPS[[j]] <- aux_list
}
GROUPS <- NEW_GROUPS

NEW_GROUPS2 <- list()
for (j in 1:length(distance)) {
  my_list <- GROUPS2[[j]]
  aux_list <- list()
  for (i in 0:(nrowY - 1)) {
    added_vectors <- lapply(my_list, function(vec) vec + i*ncolY)
    aux_list <- c(aux_list, added_vectors)
  }
  NEW_GROUPS2[[j]] <- aux_list
}
GROUPS2 <- NEW_GROUPS2

save(GROUPS, GROUPS2, file = here::here("data_files/new_groups_for_cv_new_pems_3.RData"))
```


```{r}
load(here::here("data_files/new_groups_for_cv_new_pems_3.RData"))
AUXX0 <- GROUPS2
train_test_indices_from_GROUP2 <- list()
for (i in seq_along(AUXX0)) {
  AUXX1 <- AUXX0[[i]]
  train_test_indices_from_GROUP2[[i]] <- list()
  for (j in seq_along(AUXX1)) {
    train_test_indices_from_GROUP2[[i]][[j]] <- list(train = list(AUXX1[[j]]), test = list(j))
  }
}
```


# Case $\nu = 0.5$


We first consider the stationary model.




```{r}
# Build the model
GRAPH_LME_statnu0.5 <-  graph_lme(y ~ mean_value, 
                                  graph = graph, 
                                  which_repl = 1:13, 
                                  model = list(type = "WhittleMatern", 
                                               fem = TRUE,
                                               alpha = 1,
                                               start.ltau = log_tau_from_graphlme,
                                               start.lkappa = log_kappa_from_graphlme))
summary(GRAPH_LME_statnu0.5)

POST <- posterior_crossvalidation(object = GRAPH_LME_statnu0.5, train_test_indices = train_test_indices_from_GROUP2[[1]], true_CV = FALSE)
MSE_GRAPH_LME_statnu0.5 <- POST$scores$rmse^2


save(GRAPH_LME_statnu0.5, file = "~/Desktop/GRAPH_LME_statnu0.5.RData")
```




We now fit the non-stationary model.




```{r}
# Build the model
GRAPH_LME_nonstatnu0.5 <-  graph_lme(y ~ mean_value, 
                                  graph = graph, 
                                  which_repl = 1:13, 
                                  model = list(type = "WhittleMatern", 
                                               fem = TRUE,
                                               alpha = 1,
                                               B.tau = B.tau,
                                               B.kappa =  B.kappa))
summary(GRAPH_LME_nonstatnu0.5)

POST <- posterior_crossvalidation(object = GRAPH_LME_nonstatnu0.5, mode = "loo", true_CV = FALSE)
MSE_GRAPH_LME_nonstatnu0.5 <- POST$scores$rmse^2


save(GRAPH_LME_nonstatnu0.5, file = "~/Desktop/GRAPH_LME_nonstatnu0.5.RData")
```




# Case $\nu = 1.5$


We first consider the stationary model.




```{r}
# Build the model
GRAPH_LME_statnu1.5 <-  graph_lme(y ~ mean_value, 
                                  graph = graph, 
                                  which_repl = 1:13, 
                                  model = list(type = "WhittleMatern", 
                                               fem = TRUE,
                                               alpha = 2,
                                               start.ltau = log_tau_from_graphlme,
                                               start.lkappa = log_kappa_from_graphlme))
summary(GRAPH_LME_statnu1.5)

POST <- posterior_crossvalidation(object = GRAPH_LME_statnu1.5, mode = "loo", true_CV = FALSE)
MSE_GRAPH_LME_statnu1.5 <- POST$scores$rmse^2


save(GRAPH_LME_statnu1.5, file = "~/Desktop/GRAPH_LME_statnu1.5.RData")
```


We now fit the non-stationary model.





```{r}
# Build the model
GRAPH_LME_nonstatnu1.5 <-  graph_lme(y ~ mean_value, 
                                  graph = graph, 
                                  which_repl = 1:13, 
                                  model = list(type = "WhittleMatern", 
                                               fem = TRUE,
                                               alpha = 2,
                                               B.tau = B.tau,
                                               B.kappa =  B.kappa))
summary(GRAPH_LME_nonstatnu1.5)

POST <- posterior_crossvalidation(object = GRAPH_LME_nonstatnu1.5, mode = "loo", true_CV = FALSE)
MSE_GRAPH_LME_nonstatnu1.5 <- POST$scores$rmse^2

save(GRAPH_LME_nonstatnu1.5, file = "~/Desktop/GRAPH_LME_nonstatnu1.5.RData")
```



# Case $\nu$ estimated

We first consider the stationary model.




```{r}
# Build the model
GRAPH_LME_statnuest <-  graph_lme(y ~ mean_value, 
                                  graph = graph, 
                                  which_repl = 1:13, 
                                  model = list(type = "WhittleMatern", 
                                               fem = TRUE,
                                               start.ltau = log_tau_from_graphlme,
                                               start.lkappa = log_kappa_from_graphlme))
summary(GRAPH_LME_statnuest)

POST <- posterior_crossvalidation(object = GRAPH_LME_statnuest, mode = "loo", true_CV = FALSE)
MSE_GRAPH_LME_statnuest <- POST$scores$rmse^2

save(GRAPH_LME_statnuest, file = "~/Desktop/GRAPH_LME_statnuest.RData")
```



We now fit the non-stationary model.





```{r}
# Build the model
GRAPH_LME_nonstatnuest <-  graph_lme(y ~ mean_value, 
                                  graph = graph, 
                                  which_repl = 1:13, 
                                  previous_fit = GRAPH_LME_nonstatnu1.5,
                                  model = list(type = "WhittleMatern", 
                                               fem = TRUE,
                                               B.tau = B.tau,
                                               B.kappa =  B.kappa))
summary(GRAPH_LME_nonstatnuest)

POST <- posterior_crossvalidation(object = GRAPH_LME_nonstatnuest, mode = "loo", true_CV = FALSE)
MSE_GRAPH_LME_nonstatnuest <- POST$scores$rmse^2

save(GRAPH_LME_nonstatnuest, file = "~/Desktop/GRAPH_LME_nonstatnuest.RData")
```


```{r}
list_MSE_GRAPH_LME <- list(
  MSE_GRAPH_LME_statnu0.5 = MSE_GRAPH_LME_statnu0.5,
  MSE_GRAPH_LME_nonstatnu0.5 = MSE_GRAPH_LME_nonstatnu0.5,
  MSE_GRAPH_LME_statnu1.5 = MSE_GRAPH_LME_statnu1.5,
  MSE_GRAPH_LME_nonstatnu1.5 = MSE_GRAPH_LME_nonstatnu1.5,
  MSE_GRAPH_LME_statnuest = MSE_GRAPH_LME_statnuest,
  MSE_GRAPH_LME_nonstatnuest = MSE_GRAPH_LME_nonstatnuest
)
save(list_MSE_GRAPH_LME, file = here::here("data_files/list_MSE_GRAPH_LME.RData"))
```



```{r, eval = TRUE}
load(here::here("data_files/list_MSE_GRAPH_LME.RData"))
list_MSE_GRAPH_LME
```

Below we show the MSE results for the simple linear regression `mse_loocv_lm` and the kNN regression `best_KNN_mse`.


```{r, eval = TRUE}
load(here::here("data_files/new_simple_linear_regression_results.RData"))
load(here::here("data_files/new_mse_loo_results.RData"))
#load(here::here("data_files/new_mse_cv_results.RData"))
load(here::here("data_files/new_MSE_ISOCOV.RData"))
#load(here::here("data_files/mse_CV_from_MetricGraph.RData"))
#mse_CV
#mse_CV_from_MetricGraph
MSE_LOO
data.frame(LM = mse_loocv_lm, 
           kNNdistMAT = best_KNN_mse,
           MSE_ISOCOV = MSE_ISOCOV)
```


# References


```{r, eval = TRUE}
grateful::cite_packages(output = "paragraph", out.dir = ".")
```


